---
title: "Yeast Crosslinking Data Analysis Joseph Cesare"
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
---


To begin, one needs the following files in the directory:

folder "/fasta" - contains directory of input .fasta file used in the Proteome Discoverer database search

folder "/data" - contains directory of Proteome Discoverer Data exported from the tab "crosslinks" & "xiview"

folder "/data/[method] where method corresponds to the LC-MS/MS acquisition strategy. 
method = ms2 or ms2-ms3

folder "/data/[method]/[bait] where bait corresponds to the tagged protein used in the affinity purification
bait = sap30, rxt3, sin3a, sin3b








Setup R Environment by installing packages
```{r}
# install.packages("dplyr")
# install.packages("stringr")
# install.packages("rio")
# install.packages("tidyr")
# install.packages("ggplot2")
# install.packages("snakecase")
# install.packages("tidyverse")
# install.packages("seqinr")
# install.packages("tinytex")
# install.packages("VennDiagram")
# install.packages("UpSetR")
# install.packages("ComplexUpset")
```


Setup R Environment by loading packages
```{r}
library(dplyr)
library(stringr)
library(rio)
library(tidyr)
library(ggplot2)
library(snakecase)
library(tidyverse)
library(seqinr)
library(tinytex)
library(VennDiagram)  # For Venn diagrams (for 2–3 experiments)
library(UpSetR)       # For UpSet plots (for >3 experiments)
library(ComplexUpset)
```

Begin with .fasta file processing. Here the key was extracting the GN = "" from the heading of the .fasta file. 

```{r}
#import data, show that fasta is amino acid sequence
fasta_sequences <- read.fasta(file = "fasta/Saccharomyces cerevisiae S288C 9sp_canonical TaxID=559292) 20240130 reviewed.fasta", seqtype = "AA")

#extract gene name using the GN= syntax in header
get_gene_info <- function(header) {
  gene_info <- sub(".*GN=([^ ]*).*", "\\1", header)
  return(gene_info)
}
#make gene name into a list
gene_info_list <- sapply(getAnnot(fasta_sequences), get_gene_info)
print(gene_info_list)

#extract uniprot id using the syntax in the header
get_uniprot_info <- function(header) {
  uniprot_info <- sub(".*\\|(.*?)\\|.*", "\\1", header)
  return(uniprot_info)
}

#make uniprot id into a list
uniprot_info_list <- sapply(getAnnot(fasta_sequences), get_uniprot_info)
print(uniprot_info_list)

#create dataframe from lists
id_df <- data.frame(uniprot = uniprot_info_list, gene = gene_info_list)
print(id_df)

write.csv(id_df, file = "fasta/fasta_accession_yeast_20230130.csv", row.names = FALSE)

```


Now define specific gene lists relevant to this project. For us, this is Rpd3L, Rpd3S, histones, other hdac containing complexes, other chromatin remodeling complexes

```{r}

#from this database, make subset databases for lists of genes of interest------

#rpd3large histone deacetylase complex
text <- "ASH1:CTI6:DEP1:PHO23:RPD3:RXT2:RXT3:SAP30:SDS3:SIN3:UME1:UME6"
list <- strsplit(text, ":")[[1]]
rpd3l <- id_df %>% filter(gene %in% list)
print(rpd3l)
write.csv(rpd3l, file = "fasta/yeast_rpd3l_accession.csv", row.names = FALSE)

#rpd3small histone deacetylase complex
text <- "EAF3:RCO1:RPD3:SIN3:UME1"
list <- strsplit(text, ":")[[1]]
rpd3s <- id_df %>% filter(gene %in% list)
print(rpd3s)
write.csv(rpd3s, file = "fasta/yeast_rpd3s_accession.csv", row.names = FALSE)
```

With .fasta files processed, clear the dataframe for next step
```{r}
#clear previous data from dataframe
rm(list = ls());gc()
```


Create a function to format crosslinking data, add gene names, and sort into lists based on fasta files

```{r}
# Define the function for: 
# data wrangling, formatting, add gene name, export sorted lists
process_xl_data <- function(data_path, fasta_path, fasta_list_path, output_data_gene, output_data_path, output_subset_path) {
  # Data import ---------------------------------------------------------------
  data <- import(data_path, setclass = "tibble")
  fasta <- import(fasta_path, setclass = "tibble")
  fasta_list <- import(fasta_list_path, setclass = "tibble")

  colnames(data) <- snakecase::to_any_case(colnames(data), case="snake")
    
  # Formatting input files ------------------------------------------------------
  data <- data %>% 
    rename_all(tolower) %>% 
    rename_all(~str_replace_all(., "\\W", "_")) %>%
    rename_all(~str_replace_all(., "___", "_")) %>%
    rename_all(~str_replace_all(., "__", "_")) %>%
    select(-crosslink_sequence)
  
  # Reassign uniprot with gene -------------------------------------------------
  data_gene <- data %>% rename(uniprot = accession_b)
  
  data_gene <- data_gene %>% 
    left_join(fasta, by = "uniprot") %>%
    mutate(accession_b = uniprot, 
           gene_b = ifelse(!is.na(gene), gene, uniprot)) %>%
    select(-uniprot) %>%
    select(-gene)
  
  data_gene <- data_gene %>% rename(uniprot = accession_a)
  
  data_gene <- data_gene %>% 
    left_join(fasta, by = "uniprot") %>%
    mutate(accession_a = uniprot, 
           gene_a = ifelse(!is.na(gene), gene, uniprot)) %>%
    select(-uniprot) %>%
    select(-gene)
  
  # Subset genes based on proteins of interest ---------------------------------
  data_subset <- data_gene %>%
    filter(gene_a %in% fasta_list$gene | gene_b %in% fasta_list$gene)
  
  data_subset_subset <- data_gene %>%
    filter(gene_a %in% fasta_list$gene & gene_b %in% fasta_list$gene)
  
  # Write output files
  write.csv(data_gene, file = output_data_gene, row.names = FALSE)
  write.csv(data_subset, file = output_data_path, row.names = FALSE)
  write.csv(data_subset_subset, file = output_subset_path, row.names = FALSE)
  
  # Return the processed data as a list for further use in R
  return(list(data_gene = data_gene, data_subset = data_subset, data_subset_subset = data_subset_subset))
}
```

Create function to make new directories for each specific experiment

```{r}
# Define function for:
# Create directories
create_directory <- function(folder_path) {
  if (!dir.exists(folder_path)) {  # Check if the directory already exists
    dir.create(folder_path, recursive = TRUE)  # Create the directory
    message("Directory created: ", folder_path)
  } else {
    message("Directory already exists: ", folder_path)
  }
}
```

Define major directories

```{r}
#create results/ms2 directory
if (!dir.exists("results/ms2")) {
  dir.create("results/ms2", recursive = TRUE)
}

#create results/ms2-ms3 directory
if (!dir.exists("results/ms2-ms3")) {
  dir.create("results/ms2-ms3", recursive = TRUE)
}
```


Use the functions 
data_path         refers to the crosslink data from proteome discoverer

fasta_path        refers to the fasta acession file from the database 

fasta_list_path   refers to the subset fasta file for filtering

output_data_gene  refers to output where gene name is added

output_data_path  refers to output where there must be 1 crosslink to                                           

output_subset_path    refers to output where only crosslinks within a subset 
                      genes defined by fasta_list_path pass filter


```{r}
# pd3.2 crosslinks output -------------------------------------------------

### ms2 ###

# by4741 #

#pd3.2 crosslinks output by4741ref

if (!dir.exists("results/ms2/by4741ref")) {
  dir.create("results/ms2/by4741ref", recursive = TRUE)
}

result <- process_xl_data (
  data_path = "data/ms2/by4741/xl_data_by4741_ref.csv",
  fasta_path = "fasta/fasta_accession_yeast_20230130.csv",
  fasta_list_path = "fasta/yeast_rpd3l_accession.csv",
  output_data_gene = "results/ms2/by4741ref/data_gene_by4741_ref.csv",
  output_data_path = "results/ms2/by4741ref/data_rpd3l_by4741_ref.csv",
  output_subset_path = "results/ms2/by4741ref/data_rpd3l-rpd3l_by4741_ref.csv"
)

# sap30 #

#pd3.2 crosslinks output sap30ref

if (!dir.exists("results/ms2/sap30ref")) {
  dir.create("results/ms2/sap30ref", recursive = TRUE)
}

result <- process_xl_data (
  data_path = "data/ms2/sap30/xl_data_sap30_ref.csv",
  fasta_path = "fasta/fasta_accession_yeast_20230130.csv",
  fasta_list_path = "fasta/yeast_rpd3l_accession.csv",
  output_data_gene = "results/ms2/sap30ref/data_gene_sap30_ref.csv",
  output_data_path = "results/ms2/sap30ref/data_rpd3l_sap30_ref.csv",
  output_subset_path = "results/ms2/sap30ref/data_rpd3l-rpd3l_sap30_ref.csv"
)

### ms2-ms3 ###

# by4741 #

#pd3.2 crosslinks output by4741ref

if (!dir.exists("results/ms2-ms3/by4741ref")) {
  dir.create("results/ms2-ms3/by4741ref", recursive = TRUE)
}

result <- process_xl_data (
  data_path = "data/ms2-ms3/by4741/xl_data_by4741_ref.csv",
  fasta_path = "fasta/fasta_accession_yeast_20230130.csv",
  fasta_list_path = "fasta/yeast_rpd3l_accession.csv",
  output_data_gene = "results/ms2-ms3/by4741ref/data_gene_by4741_ref.csv",
  output_data_path = "results/ms2-ms3/by4741ref/data_rpd3l_by4741_ref.csv",
  output_subset_path = "results/ms2-ms3/by4741ref/data_rpd3l-rpd3l_by4741_ref.csv"
)

# sap30 #

#pd3.2 crosslinks output sap30ref

if (!dir.exists("results/ms2-ms3/sap30ref")) {
  dir.create("results/ms2-ms3/sap30ref", recursive = TRUE)
}

result <- process_xl_data (
  data_path = "data/ms2-ms3/sap30/xl_data_sap30_ref.csv",
  fasta_path = "fasta/fasta_accession_yeast_20230130.csv",
  fasta_list_path = "fasta/yeast_rpd3l_accession.csv",
  output_data_gene = "results/ms2-ms3/sap30ref/data_gene_sap30_ref.csv",
  output_data_path = "results/ms2-ms3/sap30ref/data_rpd3l_sap30_ref.csv",
  output_subset_path = "results/ms2-ms3/sap30ref/data_rpd3l-rpd3l_sap30_ref.csv"
)

```

Create a function to also process xiview data for visualizing the subset complexes. In these files, xiview does not take gene name specifically well, so I recommend going back to the file and deleting the columns containing "gene_a" and "gene_b" for the most accurate visualization on the xiview server. I have kept gene name though because 1) it is how I filter the gene subsets 2) it allows the data to be easily readable before visualizing  

```{r}
# Define the function
process_xiview_data <- function(data_path, fasta_path, fasta_list_path, output_data_gene, output_data_path, output_subset_path) {
  # Data import ---------------------------------------------------------------
  data <- import(data_path, setclass = "tibble")
  fasta <- import(fasta_path, setclass = "tibble")
  fasta_list <- import(fasta_list_path, setclass = "tibble")
  
  
  # Formatting input files ------------------------------------------------------
  data <- data %>% 
    rename_all(tolower) %>% 
    rename_all(~str_replace_all(., "\\W", "_")) %>%
    rename_all(~str_replace_all(., "___", "_")) %>%
    rename_all(~str_replace_all(., "__", "_")) 
  
  # Reassign uniprot with gene -------------------------------------------------
  data_gene <- data %>% rename(uniprot = protein2)
  
  data_gene <- data_gene %>% 
    left_join(fasta, by = "uniprot") %>%
    mutate(protein2 = uniprot, 
           gene2 = ifelse(!is.na(gene), gene, uniprot)) %>%
    select(-uniprot) %>%
    select(-gene) 
  
  data_gene <- data_gene %>% rename(uniprot = protein1)
  
  data_gene <- data_gene %>% 
    left_join(fasta, by = "uniprot") %>%
    mutate(protein1 = uniprot, 
           gene1 = ifelse(!is.na(gene), gene, uniprot)) %>%
    select(-uniprot) %>%
    select(-gene)
  
  
  # Subset genes based on proteins of interest ---------------------------------
  data_subset <- data_gene %>%
    filter(gene1 %in% fasta_list$gene | gene2 %in% fasta_list$gene)
  
  data_subset_subset <- data_gene %>%
    filter(gene1 %in% fasta_list$gene & gene2 %in% fasta_list$gene)
  
  # Write output files
  write.csv(data_gene, file = output_data_gene, row.names = FALSE)
  write.csv(data_subset, file = output_data_path, row.names = FALSE)
  write.csv(data_subset_subset, file = output_subset_path, row.names = FALSE)
  
  # Return the processed data as a list for further use in R
  return(list(data_gene = data_gene, data_subset = data_subset, data_subset_subset = data_subset_subset))
}

```

Use the functions

```{r}
# pd3.2 xiview output -------------------------------------------------

### ms2 ###

# by4741 #

#pd3.2 xiview output by4741ref

if (!dir.exists("results/ms2/by4741ref")) {
  dir.create("results/ms2/by4741ref", recursive = TRUE)
}

result <- process_xiview_data (
  data_path = "data/ms2/by4741/xiview_by4741_ref.csv",
  fasta_path = "fasta/fasta_accession_yeast_20230130.csv",
  fasta_list_path = "fasta/yeast_rpd3l_accession.csv",
  output_data_gene = "results/ms2/by4741ref/xiview_gene_by4741_ref.csv",
  output_data_path = "results/ms2/by4741ref/xiview_rpd3l_by4741_ref.csv",
  output_subset_path = "results/ms2/by4741ref/xiview_rpd3l-rpd3l_by4741_ref.csv"
)

# sap30 #


#pd3.2 xiview output sap30ref

if (!dir.exists("results/ms2/sap30ref")) {
  dir.create("results/ms2/sap30ref", recursive = TRUE)
}

result <- process_xiview_data (
  data_path = "data/ms2/sap30/xiview_sap30_ref.csv",
  fasta_path = "fasta/fasta_accession_yeast_20230130.csv",
  fasta_list_path = "fasta/yeast_rpd3l_accession.csv",
  output_data_gene = "results/ms2/sap30ref/xiview_gene_sap30_ref.csv",
  output_data_path = "results/ms2/sap30ref/xiview_rpd3l_sap30_ref.csv",
  output_subset_path = "results/ms2/sap30ref/xiview_rpd3l-rpd3l_sap30_ref.csv"
)

### ms2-ms3 ###

# by4741 #

#pd3.2 xiview output by4741ref

if (!dir.exists("results/ms2-ms3/by4741ref")) {
  dir.create("results/ms2-ms3/by4741ref", recursive = TRUE)
}

result <- process_xiview_data (
  data_path = "data/ms2-ms3/by4741/xiview_by4741_ref.csv",
  fasta_path = "fasta/fasta_accession_yeast_20230130.csv",
  fasta_list_path = "fasta/yeast_rpd3l_accession.csv",
  output_data_gene = "results/ms2-ms3/by4741ref/xiview_gene_by4741_ref.csv",
  output_data_path = "results/ms2-ms3/by4741ref/xiview_rpd3l_by4741_ref.csv",
  output_subset_path = "results/ms2-ms3/by4741ref/xiview_rpd3l-rpd3l_by4741_ref.csv"
)

# sap30 #


#pd3.2 xiview output sap30ref

if (!dir.exists("results/ms2-ms3/sap30ref")) {
  dir.create("results/ms2-ms3/sap30ref", recursive = TRUE)
}

result <- process_xiview_data (
  data_path = "data/ms2-ms3/sap30/xiview_sap30_ref.csv",
  fasta_path = "fasta/fasta_accession_yeast_20230130.csv",
  fasta_list_path = "fasta/yeast_rpd3l_accession.csv",
  output_data_gene = "results/ms2-ms3/sap30ref/xiview_gene_sap30_ref.csv",
  output_data_path = "results/ms2-ms3/sap30ref/xiview_rpd3l_sap30_ref.csv",
  output_subset_path = "results/ms2-ms3/sap30ref/xiview_rpd3l-rpd3l_sap30_ref.csv"
)
```

With data files processed, clear the dataframe for next step
```{r}
#clear previous data from dataframe
rm(list = ls());gc()
```
